---
title: "Project 1: Boston Crime Data Analysis and Visualization"
author: "Dipti Bhattad & Shruthi Subbaiah Machimada"
date: "February 19, 2019"
output: word_document
---

Reading the csv file and subsetting the data

```{r}
crime <- read.csv("~/Desktop/Visualization/BostonCrime.csv")

crime_subset <- subset(crime, year >2015 & year <2019)

crime_shooting <- subset(crime_subset, shooting == 'Y')
```

(Since data for full years is available only for 2016-2018, disregarding other years- 2015 and 2019)

Loading the libraries

```{r}
library('ggplot2')
library('dplyr')
library('gridExtra')
```

Year-wise trend in Crimes and Shootings

```{r}
p1<-ggplot(data=summarise(group_by(crime_subset,year),incidents=n()),
       aes(year, incidents)) + scale_x_continuous(breaks=seq(2016,2018,1))+
  geom_line(color="blue")+ geom_point(size=1.5,color="dark blue") + ylab("Total # of crimes") + xlab("Year") + theme_classic() + ggtitle("Total Crimes by Year")

p2<-ggplot(data=summarise(group_by(crime_shooting,year),shooting=n()),
       aes(year, shooting)) + scale_x_continuous(breaks=seq(2016,2018,1))+scale_y_continuous(breaks=seq(50,500,25))+
  geom_line(color="orange") + geom_point(size=1.5,color="dark orange") + ylab("Total # of shootings") + xlab("Year") + theme_classic() 

grid.arrange(p1,p2)
```
2017 had a huge increase in crimes. This was observed in all major cities across the US. 

Month-wise trend in Crimes
```{r}
month_crimes<-crime_subset %>% group_by(year,month) %>% count()
month_crimes$month<-month.abb[month_crimes$month]
month_crimes$id<-rep(1:12,3)
ggplot(data=month_crimes,aes(reorder(month,id),n)) + geom_line(aes(group=as.factor(year),color=as.factor(year)), lwd = 0.75) + ylab("Total number of crimes ")+ xlab("Months") + theme_classic() + theme(legend.title=element_blank()) + ggtitle("Total Crimes by Months")
```
Feburary seems to be the month with least crimes. Looking into the mean crimes per day will help us test that.

Mean crimes per day
```{r}
x<-tidyr::separate(crime_subset,occurred_on_date,c("date","time"),sep=" ")
daily_crimes_cnt<-summarise(group_by(x,month,date,year),crime_count=n())
mean(daily_crimes_cnt$crime_count)
```

Mean crimes per day for each month

```{r}
daily_avg_by_month<-summarise(group_by(daily_crimes_cnt,year,month),daily_avg=mean(crime_count))
daily_avg_by_month$month<-month.abb[daily_avg_by_month$month]
daily_avg_by_month$id<-rep(1:12,3)
```

Plot of daily means for each month

```{r}
ggplot(data=daily_avg_by_month,
       aes(reorder(month,id), daily_avg)) +
  geom_line(aes(group=as.factor(year),color=as.factor(year)), lwd = 0.75) + ylab("Average crimes per day ") + xlab("Months") +ylim(240,310) + theme_classic() + theme(legend.title=element_blank()) + ggtitle("Mean Crimes (each day) by Months")
```
Feburary does not have the least mean crimes per day for every year. The drop in crimes in Feburary may be because Feburary has lesser number of days.
Most crimes occur between May to September. This may be because the weather is warmer in summer, or because the days are longer and most crimes occur during the day.


Hour-wise trend in Crimes

```{r}
p1 <- ggplot(data = crime_subset) + geom_histogram(mapping = aes(x=hour),fill=c(rep("#333333",6),rep("#FFFF99",6),rep("#FFCC00",6),rep("#FF6633",6)),width=1,stat="count", color = "white") +ylab("Number of crimes")+theme_classic()+scale_x_continuous(breaks=seq(0,23,1)) + ggtitle("Total Crimes by Hours")

p2 <- ggplot(data = crime_shooting) + geom_histogram(mapping = aes(x=hour),fill=c(rep("#333333",6),rep("#FFFF99",5),rep("#FFCC00",6),rep("#FF6633",6)),width=1,stat="count", color = "white") +ylab("Number of shootings")+theme_classic()+scale_x_continuous(breaks=seq(0,23,1))

grid.arrange(p1,p2)
```
The colours depict the 4 times in a day -Night, Morning, Noon and Evening. The crimes decrease after night and are least in the early hours of morning. The crimes increase again during the day.
Most crimes occur during the day.
Most shootings occur in the evening and night.


Crimes on few Main Holidays- 4th of July, Thanksgiving Day, Christmas and New Years Eve

```{r}
x<-tidyr::separate(crime_subset,occurred_on_date,c("date","time"),sep=" ")
dec25 <- subset(x,date %in% c("2018-12-25","2017-12-25","2016-12-25"))
dec31 <- subset(x,date %in% c("2018-12-31","2017-12-31","2016-12-31"))
thanksgiving <- subset(x,date %in% c("2018-11-22","2017-11-23","2016-11-24"))
independence <- subset(x,date %in% c("2018-07-04","2017-07-04","2016-07-04"))

#install.packages("directlabels")
library('directlabels')

ggplot()+
  geom_line(data=summarise(group_by(dec25,year),incidents=n()), mapping = aes(x = year, y = incidents, color = "#fe5970")) + geom_point(data=summarise(group_by(dec25,year),incidents=n()), mapping = aes(x = year, y = incidents),size=1.5, color="yellow3") +
geom_dl(data=summarise(group_by(dec25,year),incidents=n()), mapping = aes(x = year, y = incidents,label = "Dec 25"), method = "top.qp") +
  geom_line(data=summarise(group_by(dec31,year),incidents=n()), mapping = aes(x = year, y = incidents, color = "#56B4E9")) + geom_point(data=summarise(group_by(dec31,year),incidents=n()), mapping = aes(x = year, y = incidents),size=1.5, color="dark green") +
geom_dl(data=summarise(group_by(dec31,year),incidents=n()), mapping = aes(x = year, y = incidents,label = "Dec 31"), method = "smart.grid") +
  geom_line(data=summarise(group_by(thanksgiving,year),incidents=n()), mapping = aes(x = year, y = incidents, color = "#E69F00")) + geom_point(data=summarise(group_by(thanksgiving,year),incidents=n()), mapping = aes(x = year, y = incidents),size=1.5, color="dark blue") +
geom_dl(data=summarise(group_by(thanksgiving,year),incidents=n()), mapping = aes(x = year, y = incidents,label = "Thanksgiving"), method = "smart.grid") +
  geom_line(data=summarise(group_by(independence,year),incidents=n()), mapping = aes(x = year, y = incidents, color = "#999999")) + geom_point(data=summarise(group_by(independence,year),incidents=n()), mapping = aes(x = year, y = incidents),size=1.5, color="red") +
geom_dl(data=summarise(group_by(independence,year),incidents=n()), mapping = aes(x = year, y = incidents,label = "4th of July"), method = "top.qp") + scale_x_continuous(breaks=seq(2016,2018,1)) + theme_classic() + theme(legend.title=element_blank()) + ggtitle("Total Crimes on Main Holidays") + xlab("Year") + ylab("Count of Crimes")

```
On Thanksgiving and Christmas, the number of crimes is a lot less that the mean number of crimes per day for those months.

Day-wise trend in Crimes

```{r}
d <- summarise(group_by(crime_subset,day_of_week),incidents=n()) %>% arrange(desc(incidents))
ggplot(d, aes(x = reorder(day_of_week,incidents), y = incidents)) + geom_bar(stat = "identity", fill = "blue", color = "grey40", alpha = .5, width = 0.5, size = 0.5) + coord_flip() + ggtitle("Crimes by Days") + theme_classic() + xlab("Day of Week") + ylab("Count")
```
Most crime occur in Friday and the least occur on Saturday and Sunday. Weekends in Boston may be safer than weekdays.


Percentage of UCR crimes

```{r}
temp <- summarise(group_by(crime_subset,ucr_part), total_count = n())
data <- filter(temp, ucr_part %in% c("Part One", "Part Two", "Part Three"))
pct <- round(data$total_count/sum(data$total_count)*100)
lbls <- paste(data$ucr_part, pct) # add percents to labels
lbls <- paste(lbls,"%",sep="") # ad % to labels 
pie(data$total_count,labels = lbls, main="Pie Chart of Uniform Crime Reporting Codes", col = c("yellow3", "violetred1", "deepskyblue"))
```
The most common crimes are Ucr Part3 crimes, which are not very violent.


Top 5 UCR Part 1 crimes

### UCR Part 1 crimes
```{r}
ucr_part1 <- subset(crime_subset, ucr_part == "Part One")
```

### Top crimes in ucr part 1
```{r}
top_crimes_ucrpart1 <- ucr_part1 %>% count(offense_code_group) %>% arrange(desc(n)) %>% top_n(5)
top_crimes_ucrpart1_v <- as.character(top_crimes_ucrpart1$offense_code_group)
top_crimes_ucrpart1$Crime_Count <- top_crimes_ucrpart1$n
```

### Treemaps for most common ucr part 1 crimes
```{r}
library(treemapify)
ggplot(top_crimes_ucrpart1, aes(area=Crime_Count,label=offense_code_group,fill=Crime_Count))+geom_treemap()+ geom_treemap_text(fontface = "italic", colour = "black", place = "centre",grow = FALSE)+ggtitle("Most Common UCR Part One crimes") + scale_fill_gradient(low = "yellow", high = "dark orange")
```

Year-wise Trend in Top 5 UCR Part One Crimes

```{r}
df1 <- subset(ucr_part1, offense_code_group %in% top_crimes_ucrpart1_v)
ggplot(data = summarise(group_by(df1,offense_code_group,year), crime_count_ucr1=n())) + geom_line(aes(year,crime_count_ucr1, color = offense_code_group), lwd = 0.75) + geom_point(aes(year,crime_count_ucr1),size=1,color="black") + xlab("Year") + ylab("Total Number of Crimes") + scale_x_continuous(breaks=seq(2016,2018,1)) + theme_classic() + ggtitle("Trend of Top 5 UCR Part One Crimes by Year")
```


Distribution of Top 5 UCR Part 1 Crimes Across Top 5 Unsafest Districts

### Unsafe districts in ucr part 1
```{r}
unsafe_dist_ucr1 <- ucr_part1 %>% count(district) %>% arrange(desc(n)) %>% top_n(5)
unsafe_dist_ucr1_v <- as.character(unsafe_dist_ucr1$district)
```

### Heatmap for all years together
```{r}
heatmap_data = subset(ucr_part1, offense_code_group %in% top_crimes_ucrpart1_v & district %in% unsafe_dist_ucr1_v)
ggplot(data = summarise(group_by(heatmap_data,district,offense_code_group), cc = n())) + geom_tile(aes(x=district,y=offense_code_group, fill=cc)) + scale_fill_gradient2(low = "beige", mid = "orange", high = "brown", midpoint = 3500) + ggtitle("Heatmap for all years") + xlab("Districts") + ylab("Crimes")

```
Larceny, the most common UCR Part 1 crime is highly distributed across all districts. The next common crime is Larceny from Motor Vehicle, which is not that high.
The distribution of these crimes were similar for all years, so we aggregated all years in one heatmap.


Map of Boston districts and the distribution of crime across them
```{r}
library(maptools)
#area<-readShapePoly("Boston_Neighborhoods.shp") 
#area_df<-fortify(area)
area_df<-read.csv("~/Desktop/Visualization/Area_Maps.csv")
area_df$safety<-3
#Most crimes
#Roxbury
area_df <- within(area_df, safety[group == 8.1] <- 0) #B2
area_df <- within(area_df, safety[group == 2.1] <- 0) #B2
area_df <- within(area_df, safety[group == 3.1] <- 0) #B2
#Dorchester
area_df <- within(area_df, safety[group == 21.1] <- 0.7) #C11
#Fenway,Backbay
area_df<-within(area_df, safety[group == 16.1] <- 1) #D4
#Mattapan
area_df<-within(area_df, safety[group == 20.1] <- 1.5) #B3
#Least crimes
area_df<-within(area_df, safety[group == 12.1] <- 6) #A15
area_df<-within(area_df, safety[group == 11.1] <- 5) #A7
area_df<-within(area_df, safety[group == 18.1] <- 4.5) #E5
area_df<-within(area_df, safety[group == 19.1] <- 4) #E18
ggplot(area_df,aes(x=long,y=lat,group=group,fill=safety))+geom_polygon(color="black")+theme_void()+scale_fill_gradient(low = "red", high = "yellow3")
```
Most populated- Dorchester(Red) and Roxbury (Dark Red)
Most number of crimes- Roxbury (Dark Red) and Dorchester(Red)
Most number of shootings - Roxbury (Dark Red) and Mattapan(light red)
Most number of UCR Part 1 crimes - Fenway, BackBay and Downtown (Orange)
Safest areas (Least crimes) - East Boston, Charleston (Light green)


Has the safety of districts changed over the years

```{r}
unsafe_dist <- crime_subset %>% count(district) %>% arrange(desc(n)) %>% top_n(5)
unsafe_dist
unsafe_dist_v <- as.character(unsafe_dist$district)
```
```{r}
df2 <- subset(crime_subset, district %in% unsafe_dist_v)
p1<-ggplot(data = summarise(group_by(df2,district,year), crime_count_dist=n())) + geom_line(aes(year,crime_count_dist, color = district)) + scale_x_continuous(breaks=seq(2016,2018,1))+ylab("Crime count")+theme_classic()

df_ucr<-subset(ucr_part1,district %in% unsafe_dist_v)
p2<-ggplot(data = summarise(group_by(df_ucr,district,year), crime_count_dist=n())) + geom_line(aes(year,crime_count_dist, color = district)) + scale_x_continuous(breaks=seq(2016,2018,1))+ylab("Ucr Part1 crimes count")+theme_classic()

grid.arrange(p1,p2)
```

The crimes in Roxbury and Dorchester have decreased in 2018, when compared to 2016 and 2017. But the UCR Part 1 crimes in these
districts have remained the same.
Given more historic data we would be able to find trends.

Unsafe Streets in Boston
```{r}
unsafe_streets = crime_subset %>% count(street) %>% arrange(desc(n)) %>% top_n(11)
unsafe_streets <- unsafe_streets[c(1,3:11),]
unsafe_streets$Crime_Count <- unsafe_streets$n
#unsafe_streets
```


Unsafe streets in Dorchester and Roxbury
```{r}
unsafe_street_RoxDorch = filter(crime_subset,district=="B2" |district=="C11" ) %>% count(street) %>% arrange(desc(n)) %>% top_n(11)
unsafe_street_RoxDorch <- unsafe_street_RoxDorch[c(1:3,5:11),]
unsafe_street_RoxDorch$Crime_Count <- unsafe_street_RoxDorch$n
#unsafe_street_RoxDorch
```

Unsafe Streets(Shooting) in Boston
```{r}
unsafe_street_shooting = subset(crime_shooting) %>% count(street) %>% arrange(desc(n)) %>% top_n(11)
unsafe_streets_shooting <- unsafe_street_shooting[c(1:4),]
unsafe_streets_shooting$Shooting_Count <- unsafe_streets_shooting$n

```

Unsafe streets in boston, and Unsafe streets in Roxbury and Dorchester
```{r}
t1 <- ggplot(unsafe_streets, aes(area=Crime_Count,label=street,fill=Crime_Count))+geom_treemap()+ geom_treemap_text(fontface = "italic", colour = "black", place = "centre",grow = FALSE)+ggtitle("Tree Maps for Unsafest Streets") + scale_fill_gradient(low = "yellow3", high = "red")

t2 <- ggplot(unsafe_street_RoxDorch, aes(area=Crime_Count,label=street,fill=Crime_Count))+geom_treemap()+ geom_treemap_text(fontface = "italic", colour = "black", place = "centre",grow = FALSE)+ggtitle("Unsafe Streets in Roxbury and Dorchester") + scale_fill_gradient(low = "light pink", high = "orange")

t3 <- ggplot(unsafe_streets_shooting, aes(area=Shooting_Count,label=street,fill=Shooting_Count))+geom_treemap()+ geom_treemap_text(fontface = "italic", colour = "black", place = "centre",grow = FALSE)+ggtitle("Unsafe Streets- Shooting") + scale_fill_gradient(low = "beige", high = "brown")

grid.arrange(t1,t2,t3)
```


Daily pattern of crimes on the unsafe streets
```{r}
unsafe_streets_trend<- crime_subset[crime_subset$street %in% unsafe_streets$street,]
ggplot(data=summarise(group_by(unsafe_streets_trend,hour,street),crime_count=n()))+geom_line(aes(hour,crime_count,color=street))+scale_x_continuous(breaks=0:23,1)+xlab("Hour of day")+theme_classic()
```
Crimes increase in all these streets during the day, with most crimes accuring from 5 PM to 8 PM.
Crimes in Boylston street and Washington street follow a very unpredictable pattern.


Crimes which involved shooting
```{r}
shooting_offenses<-crime_shooting %>% count(offense_code_group) %>% arrange(desc(n)) %>% top_n(10)
```

```{r}
ggplot(data=shooting_offenses,aes(x=reorder(offense_code_group,n),y=n))+geom_col(aes(fill=offense_code_group))+coord_flip()+scale_x_discrete(breaks=seq(0,600,100))+
  xlab("Offense")+ylab("Number of shootings")+theme_classic()
```












